---
title: "Machine learning species distribution models in R: from zero to pro(blems)"
author: "Victor Felix"
date: "2026-01-05"
output:
  pdf_document: default
  html_document: default
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

Let's load some libraries

```{r}
#install.packages("pROC") # if not installed
library(pROC)
#install.packages("gbm") # if not installed
library(gbm)
#install.packages("dismo") # if not installed
library(dismo)
#install.packages("reshape2") # if not installed
library(reshape2)
#install.packages("dplyr") # if not installed
library(dplyr)
#install.packages("geodata") # if not installed
library(geodata)
#install.packages("terra") # if not installed
library(terra)
#install.packages("tigris") # if not installed
library(tigris)
#install.packages("sf") # if not installed
library(sf)
#install.packages("ggplot2") # if not installed
library(ggplot2)
#install.packages("caret")   # run once
library(caret)
```

Here I get the shapefiles of each country in our study region which I will use for data selection later.

```{r}
# US shapefiles from tigris
us_states <- tigris::states(year = 2023, cb = TRUE)
# plot(us_states) # inspect data

# Selecting only the contiguous US
contiguous_us <- us_states[!us_states$STUSPS %in% c("AK", "HI", "PR", "GU", "VI", "MP", "AS"), ]

# Combine the states on the contiguous US
contiguous_us_sf <- sf::st_union(contiguous_us)

# Transform into a vector - not an sf object
contiguous_us_vect <- terra::vect(contiguous_us_sf)

# Plot contigous US
png("figures/contiguous_us_map.png", width = 1000, height = 800, res = 250)
plot(contiguous_us_vect, main="Contiguous United States")
dev.off()

plot(contiguous_us_vect)
```

# Downloading WorldClim rasters

```{r}
# Downloading WorldClim rasters
worldclim_rasters<-geodata::worldclim_global(var = "bio", # variables to download
                                 res = 5, # choose the resolution
                                 path = "data/") # path were data is saved
plot(worldclim_rasters)
```
Download WorldClim data.

https://www.worldclim.org/data/bioclim.html#google_vignette

```{r}
# New simplified names
new_names <- c(
  "AnnMeanTemp", "MeanDiurnalRange", "Isotherm", "TempSeasonality",
  "MaxTempWarmM", "MinTempColdM", "TempAnnRange", "MeanTempWetQ",
  "MeanTempDryQ", "MeanTempWarmQ", "MeanTempColdQ", "AnnPrecip",
  "PrecipWetM", "PrecipDryM", "PrecipSeasonality", "PrecipWetQ",
  "PrecipDryQ", "PrecipWarmQ", "PrecipColdQ"
)

# Rename rasters
base::names(worldclim_rasters) <- new_names

plot(worldclim_rasters)

```
Crop the stacked raster to the size of the contigous US and rename for easier interpretation.

```{r}
# Define coordinates for contiguous US
us_extent <- ext(-125, -66.5, 24.5, 49.5)  # approximate lat/lon 

# Crop the rasters
worldclim_rasters_us <- terra::crop(worldclim_rasters, contiguous_us_vect)%>%
  mask(contiguous_us_vect)

# save
png("figures/worldclim_rasters.png", width = 1000, height = 600, res = 150)
plot(worldclim_rasters_us)
dev.off()

# plot
plot(worldclim_rasters_us)
```

Download the American green tree frog data.

https://www.gbif.org/species/10810025

Dryophytes cinereus (Schneider, 1799)

GBIF.org (11 December 2025) GBIF Occurrence Download https://doi.org/10.15468/dl.kg3eax

```{r }
# Loading the dataset

AGTF_raw <- readr::read_delim("data/american_green_tree_frog.csv", delim = "\t", escape_double = FALSE, 
    col_names = TRUE, na = c("", "NA"))

head(AGTF_raw)
```

Getting coordinates from the GIBF Vespa Velutina data.

```{r}
# Create a data frame using latitudes and longitudes
AGTF_occurences <- base::data.frame("latitude" = AGTF_raw$decimalLatitude, 
                            "longitude" = AGTF_raw$decimalLongitude,
                            "id"=1)
# NROW(AGTF_raw_data)  
  
# Remove NAs - places without Latitude or Longitude data
AGTF_occurences <- stats::na.omit(AGTF_occurences)
# NROW(AGTF_raw_data) 

# Create a vector with these locations (shapefile points)
AGTF_occurences <- terra::vect(AGTF_occurences, geom = c("longitude", "latitude"))

# Plotting data and contiguous US
plot(AGTF_occurences, main="AGTF occurences")
plot(contiguous_us_vect, add=TRUE)
```
We will use the contiguous US vector to select only occurrences in the study region. 

!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
# this step take some time to run  
# go to the slides to talk about next step
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

```{r}
# Crop the raster using the shapefile's extent
AGTF_occurences_US <- terra::crop(AGTF_occurences, contiguous_us_vect)

# save
png("figures/occurrences_points.png", width = 1000, height = 600, res = 150)
plot(AGTF_occurences_US, main="Green Treefrog occurrences points")
plot(contiguous_us_vect ,add=TRUE)
dev.off()

```
Transforming the points (vector format) to raster

```{r}
# Computing the number of occurrences in each raster cell.
AGTF_occurences_raster <- terra::rasterize(AGTF_occurences_US, field="id", worldclim_rasters_us$AnnMeanTemp, fun = "count", background = 0, filename="data/AGTF_count.tif", overwrite=TRUE)

# save
png("figures/occurrences_raster.png", width = 1000, height = 600, res = 150)
plot(AGTF_occurences_raster)
plot(contiguous_us_vect, border = "white",add=TRUE, main="Green Treefrog raster", sub = "Number of occurences per raster cell")
dev.off()

# Plot raster
plot(AGTF_occurences_raster)
plot(contiguous_us_vect, border = "white",add=TRUE, main="American Green Treefrog raster", sub = "Number of occurences per raster pixel")

```
Plotting with ggplot for better visualization of occurrences 

```{r}
# Convert raster to data frame to plot with ggplot
AGTF_occurences_df <- base::as.data.frame(AGTF_occurences_raster, xy = TRUE)
colnames(AGTF_occurences_df) <- c("x", "y", "value")

# Filter dataframe to keep only values > 0
AGTF_occurences_df <- AGTF_occurences_df[AGTF_occurences_df$value > 0, ]

AGTF_occurences_plot<-ggplot() +
  geom_sf(data = contiguous_us_sf, fill = "white", color = "black", linewidth = 0.2) +
  geom_point(
    data = AGTF_occurences_df,
    aes(x = x, y = y, size = value, color = value),
    alpha = 0.6
  ) +
  scale_size_continuous(
  range = c(1, 10),
  name = "",
  breaks = c(1, 50, 100, 200, 300)
) +
  scale_color_gradient(low = "orange", high = "darkred", name = "") +
  labs(
    title = "Green Tree Frog occurrences",
    subtitle = "Number of occurences per raster pixel",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 19),
    plot.subtitle = element_text(size = 14),
    legend.position = "bottom"
  )

AGTF_occurences_plot

ggsave("figures/density_plot.jpg", AGTF_occurences_plot, width = 8,
  height = 6,  dpi = 300)
```
Let's say that at least one occurrence per pixel represent presence

```{r}

# Set pixels values with at least one occurrence to 1 (presence)
AGTF_occurences_final_raster <- AGTF_occurences_raster
AGTF_occurences_final_raster[AGTF_occurences_final_raster > 0] <- 1

# Plot for visualization
plot(AGTF_occurences_final_raster, col = c("white", "black"))
plot(contiguous_us_vect,add=TRUE, main="Presence points raster", sub = "1 means presence or at least one occurrence per pixel")

```
Now we will build a dataframe with coordinates of the presence points.

```{r}
# Convert raster to data frame
AGTF_occurences_final_df <- base::as.data.frame(AGTF_occurences_final_raster, xy = TRUE)
colnames(AGTF_occurences_final_df) <- c("longitude", "latitude", "value")

AGTF_presence_df <- AGTF_occurences_final_df %>%
  filter(value == 1)

# Save the plot
png("figures/presence_points.png", width = 1000, height = 600, res = 150)
plot(contiguous_us_vect,
     xlab = "Longitude", ylab = "Latitude",
     main = "Presence points")
points(AGTF_presence_df$longitude, AGTF_presence_df$latitude,
       pch = 16, cex = 0.2, col = "darkgreen")
dev.off()

# Plot presence points
plot(contiguous_us_vect,
     xlab = "Longitude", ylab = "Latitude",
     main = "Presence points")
points(AGTF_presence_df$longitude, AGTF_presence_df$latitude,
       pch = 16, cex = 0.3, col = "darkgreen")
```
######
Back to slides
#####

Now we will create the dataframe with climate features for the presence point and the background points their climate features.


```{r}
AGTF_presence_points <- terra::vect(
  AGTF_presence_df,
  geom = c("longitude", "latitude")
)

presence_points <- terra::extract(worldclim_rasters_us, AGTF_presence_points)

# Drop ID column
presence_points <- base::subset(presence_points, select = -c(ID))
# Create a Label column with 1 as presence
presence_points$Label <- 1

presence_points
```

Now we randomly select the same number of background points as presence points, and sample the climate data from each of these points.

```{r}
set.seed(30)
# Sampling random background points 
x_y_background_points <- spatSample(worldclim_rasters_us, size = NROW(presence_points), method = "random", na.rm = TRUE, xy=TRUE)%>%
  mutate(Label = 0)

# save the plot
png("figures/background_points.png", width = 1000, height = 600, res = 150)
plot(x_y_background_points$x, x_y_background_points$y,
     pch = 16, cex = 0.3,
     xlab = "Longitude", ylab = "Latitude",
     main = "Background points")
plot(contiguous_us_vect,add=TRUE)
dev.off()

#Plot 
plot(x_y_background_points$x, x_y_background_points$y,
     pch = 16, cex = 0.4,
     xlab = "Longitude", ylab = "Latitude",
     main = "Background points")
plot(contiguous_us_vect,add=TRUE)

# sample climate data from these points
background_points <- subset(x_y_background_points, select = -c(x, y))

head(background_points)
```

Get 50% of the data for training and 50% for testing.

```{r}
# Randomly sample 70% of the background indices
sample_indices <- base::sample(seq_len(nrow(background_points)), size = 0.7 * nrow(background_points))
# Create a sampled dataframe with 70% of the data
bg_70 <- background_points[sample_indices, ]

# Randomly sample 50% of the presence row indices
sample_indices <- base::sample(seq_len(nrow(presence_points)), size = 0.7 * nrow(presence_points))
# Create a sampled dataframe with 70% of the data
presence_70 <- presence_points[sample_indices, ]

#rbind 50% of presence and background for training data
train.data.70<-base::rbind(presence_70,bg_70)

# Get the 50% data left for presence and background data 
test_presence_30 <- presence_points[-sample_indices, ]
test_bg_30 <- background_points[-sample_indices, ]

#rbind the 50% left of presence and background for testing
test.data.30<-rbind(test_bg_30,test_presence_30)

```

Remember to talk that GBM are ihume to colinearity

```{r}

# Calculate the correlation matrix, removing NAs
corr_matrix <- cor(train.data.70[,1:20], use = "complete.obs")

#heatmap(corr_matrix, symm = TRUE)
# Melt the correlation matrix for ggplot2
melted_corr <- melt(corr_matrix)

# Create the heatmap
ggplot(data = melted_corr, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed()

```
Remove some columns I don't want to train

```{r}
############
set.seed(11)

# training data is 50% of the downsampled presence points
train.data <- train.data.70

#################
# GBM model setup
#################

train.data$Label <- as.factor(train.data$Label) # Define target column as a factor
levels(train.data$Label) <- c("Background", "Presence") # Setup levels of the target column

# Define a tuning grid for model optimization
tune_grid <- expand.grid(
  interaction.depth = seq(3,4,1), # tree depth for complex relationships
  n.trees = c(360), # number of trees
  shrinkage = c(0.01), # small learning rate
  n.minobsinnode = c(50) # leaf size - minimum number of observation that need to be in each leaf so the split exist
)

############
set.seed(11)
# Create a defined set of folds for cross-validation training
# and for reproducibility
folds <- createFolds(train.data[[20]], k = 3, returnTrain = TRUE)

############
set.seed(11)
# setup the cross-validation approach
train_control <- trainControl(
  method = "cv",  number = 3,
  verboseIter = FALSE, classProbs = TRUE,
  summaryFunction = twoClassSummary,
  index = folds  # Use reproduceble folds for cross-validation
)

############
set.seed(11)
# Train the GBM model
gbm_model <- train(
  x = train.data[, 1:19],    # predictors columns
  y = train.data[[20]],      # target column
  method = "gbm",         
  trControl = train_control, # cross-validation settings
  tuneGrid = tune_grid,      # tuning grid
  metric = "ROC"             # ROC metric
)

# Print the mean cross-validation AUC
cat("AUC:", mean(gbm_model$resample$ROC), "\n")

```

Plotting GBM training ROC

```{r}
# Get predicted probabilities 
pred_probs <- predict(gbm_model, newdata = train.data[, 1:20], type = "prob")[, "Presence"]

# Compute ROC curve
roc_obj <- roc(response = train.data$Label, predictor = pred_probs)

# Create a data frame for plotting
roc_df <- data.frame(
  tpr = roc_obj$sensitivities,
  fpr = 1 - roc_obj$specificities
)

# Plot 
train_plot <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = paste("a) GBM Training ROC"),
    subtitle = paste("AUC =", round(mean(gbm_model$resample$ROC), 3)),
    x = "Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal(base_size = 14)

#####
ggsave("figures/train_ROC.jpg", plot = train_plot, width = 3, height = 3, dpi = 300)

print(train_plot)
```

GBM testing ROC

```{r}
# Loading test data and transforming labels to factor
test.data <- test.data.30
test.data$Label <- as.factor(test.data$Label)
levels(test.data$Label) <- c("Background", "Presence")

# Forecasts for test data using our trained model
predictions <- predict(gbm_model, test.data[, 1:20], n.trees = 100, type = "prob")

# Extract probabilities
prob_positive <- predictions$Presence

# Run the ROC curve calculation
roc_curve <- roc(test.data$Label, prob_positive)

# Calculate the AUC
auc_value <- auc(roc_curve)

# Create dataframe for plotting
roc_df <- data.frame(
  fpr = 1 - roc_curve$specificities,
  tpr = roc_curve$sensitivities
)

# Plot
test_plot <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = paste("b) GBM Testing ROC"),
    subtitle = paste("AUC =", round(auc(roc_curve), 3)),
    x = "Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal(base_size = 14)

test_plot

# Save the ggplot
ggsave("figures/test_ROC.jpg", plot = test_plot, width = 3, height = 3, dpi = 300)
```

Plotting GBM variable importance.

```{r}
set.seed(11)
# Calculate variable importance
importance <- varImp(gbm_model)
# Extract the importance values as a data frame
feature_importance <- as.data.frame(importance$importance)
# Add a column for the percentage importance
feature_importance$Percentage <- (feature_importance$Overall / sum(feature_importance$Overall)) * 100
# Sort by percentage importance 
feature_importance <- feature_importance[order(-feature_importance$Percentage), ]
# Convert rownames to a column for easier plotting
feature_importance$Variable <- rownames(feature_importance)

# Plot
importance_plot <- ggplot(feature_importance, aes(x = reorder(Variable, Percentage), y = Percentage)) +
  geom_segment(aes(xend = Variable, y = 0, yend = Percentage), color = "blue") +
  geom_point(color = "blue", size = 3) +
  coord_flip() +
  labs(
    title = "GBM Feature importance",
    x = "Model features",
    y = "Percentage (%)"
  ) +
  theme_minimal()

importance_plot

```

Forecasting the contiguous U.S.

```{r}

map_pred <- terra::predict(object=worldclim_rasters_us, model=gbm_model,filename= "forecasts/US_SDM.tif",overwrite=TRUE,type = "prob")

# load the forecasts
model_pred<-rast("forecasts/US_SDM.tif")
# mask the forecasts
model_pred_masked<-mask(model_pred,contiguous_us_vect)

plot(model_pred_masked$Presence, main="Contiguos United States", xlab="",
     col = viridis::viridis(100),
     range = c(0, 1))
plot(contiguous_us_vect, add=TRUE)


```
Forecasting the whole world

```{r}

map_pred <- terra::predict(object=worldclim_rasters, model=gbm_model,filename= "forecasts/WORLD_SDM.tif",overwrite=TRUE,type = "prob")

# load the forecasts
model_pred<-rast("forecasts/WORLD_SDM.tif")

# plot
plot(model_pred$Presence, main="World", xlab="",
     col = viridis::viridis(100),
     range = c(0, 1))
```